---
title: "607 Assignment 2"
author: "AJ Strauman-Scott"
output: html_document
date: "2024-02-03"
---
Because we were unable to successfully install and set-up the PostgreSQL and ODBC drivers for RStudio that would allow us to make a direct connection to the databases in PostgreSQL, we have redone this homework to export the tables we made in PostgreSQL to.csv, and we will now import those files from where I have uploaded them on GitHub.

I have uploaded my issues to our Slack channel for help and will continue to troubleshoot the issue.


```{r}
# Import the data
movies <- data.frame(read.csv("https://github.com/annejscott/DATA607/raw/main/607_assignment_2/exported_from_sql/wk2assignments.movies.csv", header = FALSE, sep = ","))

moviescol <- c("id", "title", "year", "actor", "actress", "imdb")

colnames(movies) <- moviescol

head(movies)
```

```{r}
ratings <- data.frame(read.csv("https://github.com/annejscott/DATA607/raw/main/607_assignment_2/exported_from_sql/wk2assignments.ratings.csv", header = FALSE, sep = ","))

ratingscol <- c("person_id", "movie_id", "rating")

colnames(ratings) <- ratingscol

head(ratings)
```

```{r}
surveytakers <- data.frame(read.csv("https://github.com/annejscott/DATA607/raw/main/607_assignment_2/exported_from_sql/wk2assignments.surveytakers.csv", header = FALSE, sep = ","))

peoplecol <- c("id", "name")

colnames(surveytakers) <- peoplecol
head(surveytakers)
```