---
title: "Lab 1-1 - Importing and Subsetting a Dataset"
author: "AJ Strauman-Scott"
output: html_document
date: "2024-02-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, import libraries to use in the session. Usually this code would be hidden
```{r}
library('tidyverse')
```

Next, import the data into the RStudio project as a data frame. We are using the `na.strings` aruegment to replace missing values coded as "?" strings to missing values recognized by the program.

```{r}
bridges <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/bridges/bridges.data.version2", header= FALSE, sep=",",na.strings = "?")
```

Give the columns names based on the data director

```{r}
colnames(bridges) <- c("ID", "RIVER", "LOCATION","ERECTED","PURPOSE","LENGTH","LANES","CLEAR-G","T-OR-D","MATERIAL","SPAN","REL-L","TYPE")
```

Subset the data to focus on only those observations that interest us. For instance, how many of the aqueducts are made of wood?

```{r}
wood_aqueducts <- subset(bridges, PURPOSE == "AQUEDUCT" & MATERIAL == "WOOD")
head(wood_aqueducts)
```

There are only 3 wooden aqueducts in Pittsburgh, it seems!
