---
title: "Data 607 Project 1"
author: "AJ Strauman-Scott"
date: "2024-02-12"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

# Introduction

For this project, we have received a text file with chess tournament results. This text file, while legible to the human eye, is in a format that is very difficult for a computer to read.

Our goal is to input and clean the data, calculate a new variable to display the average pre-tournament ranking for each player's opponents across the 7 rounds, and then export a .csv file containing a data table with the player's names, states, total points, pre-tournament ranking, and average pre-tournament ranking of their opponent.

#### Collaboration

The author worked in a group with Evelyn, James and Dhanya on Slack to troubleshoot through this project.

#### Load Libraries

We will use the `tidyverse`, `knitr` and `rebus` packages.

```{r load libraries, message = FALSE}
library(tidyverse)
library(knitr)
library(rebus)
```

## Importing the Data

We begin by importing the .txt file from Github. We specify the deliminator as the bar `|`, that there are no column names, and to skip the first 4 lines of the text.

```{r import data, warning = FALSE}
# Import the Files
chessImported <- read_delim("https://github.com/annejscott/DATA607/raw/main/project_1/tournamentinfo.txt", delim = "|", col_names = FALSE, skip = 4, show_col_types = FALSE)

# Convert the file to a data frame
chessImported <- data.frame(chessImported)

glimpse(chessImported)

```

As you can see, there is a limit to what our `read_delim` function can do with this file.

## Tidy The Data

Our next step is to tidy the data into a usable form that both we and the computer can read. We need to:

-   remove the empty column at the end

-   remove the rows with only dashes in the first column

-   combine rows in twos to create the full record of each observation

-   extract the relevant numerical data from the generated strings in the columns for the pre-tournament ranking and the seven rounds, and format it to the correct data type

```{r tidy the data}
# Remove the NA rows
chessImported <- subset(chessImported, chessImported$X1 != chessImported$X1[3])

# Remove the NA column
chessImported <- subset(chessImported, select = -X11)

# Separate out the data interatively
# initialize the two data frames
name_rows <- data.frame()
state_rows <- data.frame()

# Use recursion to create two data frames
for (i in 1:nrow(chessImported)) {
  this_row <- chessImported[i,]
  if (i %% 2 == 1) {
    name_rows <- rbind(name_rows, this_row)
  } 
  else {
    state_rows <- rbind(state_rows, this_row)
  }
}

# Combine the data sets
chessGoodRows <- cbind(name_rows, state_rows)

# Rename the columns
new_col_names <- c("num", "name", "points", "rd1", "rd2", "rd3", "rd4", "rd5", "rd6", "rd7", "state", "prerank")
colnames(chessGoodRows) <- new_col_names

# Keep only columns of interest and rearrange
chessRelevent <- chessGoodRows[, c("num", "name", "state","points", "prerank", "rd1", "rd2", "rd3", "rd4", "rd5", "rd6", "rd7")]

# Renumber the rows
new_row_nums <- c(1:nrow(chessRelevent))
rownames(chessRelevent) <- as.character(new_row_nums)

# Extract numeric data from strings in each of the columns
chessClean <- chessRelevent %>% 
  mutate(prerank = as.numeric(str_sub(prerank, 16, 19)), 
        rd1 = as.numeric(str_extract(rd1, "\\d+")),
        rd2 = as.numeric(str_extract(rd2, "\\d+")),
        rd3 = as.numeric(str_extract(rd3, "\\d+")),
        rd4 = as.numeric(str_extract(rd4, "\\d+")),
        rd5 = as.numeric(str_extract(rd5, "\\d+")),
        rd6 = as.numeric(str_extract(rd6, "\\d+")),
        rd7 = as.numeric(str_extract(rd7, "\\d+")),
        points = as.numeric(str_trim(points)),
        num = as.numeric(num),
        name = str_trim(name),
        state = str_trim(state)
        )

glimpse(chessClean)
```

Our data is now clean and usable for both ourselves and the computer.

## Calculate Average Opponents' Pre-Tournament Ranking

In order to output the requested variables in our output file, we need to create a new variable, that calculates the average pre-tournament ranking of each player's opponents.

To accomplish this, we use the number of each player's opponent in each round to look up that players pre-tournament ranking, and then replace the opponent's number with the opponent's pre-tournament ranking in that round's column.

From there, we calculate the mean across rows to generate our new variable.

```{r}
# Reassign prerank values to their opponent numbers in the rounds columns
rounds <- c("rd1", "rd2", "rd3", "rd4", "rd5", "rd6", "rd7")

for (rds in rounds) {
  for (opp in seq_along(chessClean[[rds]])) {
    chessClean[[rds]][opp] <- chessClean$prerank[chessClean[[rds]][opp]]
  }
}

# Calculate the average of each opponent's prerank.
chessAvg <- chessClean %>% 
  rowwise() %>%  
  mutate(opp_avg = round(mean(c(rd1, rd2, rd3, rd4, rd5, rd6, rd7), na.rm = T), digits = 0))

# Extract only columns of interest
chessFinal <- chessAvg[, c("name", "state","points", "prerank", "opp_avg")]

glimpse(chessFinal)
```

Our dataset is complete! We have all our data defined for the correct variables, in the correct data type. We're ready to export.

## Export to .csv

Now we use the `write_csv` function to export our clean and ready dataset as the requested .csv file.

```{r export.csv}
# Export .csv file
write_csv(chessFinal, "chess_tournament_final.csv")
```
